---
title: "XGBoost 30-Day Readmission Risk Healthguard"
author: "Jessica Martin"
date: "2026-09-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Importing Packages

```{r, message=FALSE, warning=FALSE}
library("xgboost")
library("tidyverse")
```
### Importing our data

```{r}
rra_df <- read_csv("readmissions_cleaned.csv")
```
### Some quick variable plots to describe the categorical data

```{r}
base_plot = ggplot(rra_df)

base_plot + geom_bar(aes(x=gender, fill = gender)) +
  theme_minimal() +
  scale_fill_manual("legend", 
        values = c("male" = "#76B7B2", "female" = "#EDC948", "other" = "#76B7B2")
        ) +
  labs(title = "Gender Distribution",
       x = "Gender",
       y = "Patient Count")
base_plot + geom_bar(aes(x=insurance_type, fill = insurance_type)) +
  theme_minimal() +
  scale_fill_manual("legend", 
        values = c("self-pay" = "#76B7B2", "private" = "#EDC948", 
                   "medicare" = "#EDC948", "medicaid" = "#76B7B2")
        ) +
  labs(title = "Insurance Type",
       x = "Insurance Type",
       y = "Patient Count")
base_plot + geom_bar(aes(x=admission_type, fill = admission_type)) +
  theme_minimal() +
  scale_fill_manual("legend",
        values = c("urgent" = "#76B7B2","elective" = "#76B7B2", "emergency" = "#EDC948")
        ) +
  labs(title = "Admission Type",
       x = "Admission Type",
       y = "Patient Count")
base_plot + geom_bar(aes(x=discharge_disposition, fill = discharge_disposition)) +
  theme_minimal() +
  scale_fill_manual("legend",
                    values = c("home" = "#EDC948", "transfer" = "#76B7B2", 
                               "rehabilitation" = "#76B7B2", "expired" = "#76B7B2")
                    ) +
  labs(title = "Discharge Disposition",
       x = "Discharge Disposition",
       y = "Patient Count")
```
### Converting our categories to factors

```{r}
rra_df$gender <- factor(rra_df$gender)
rra_df$insurance_type <- factor(rra_df$insurance_type)
rra_df$admission_type <- factor(rra_df$admission_type)
rra_df$discharge_disposition <- factor(rra_df$discharge_disposition)
rra_df$primary_diagnosis_code <- factor(rra_df$primary_diagnosis_code)
```

### Dropping duplicative columns and index columns

```{r}
rra_df <- rra_df %>% select(-patient_id, -diagnosis_desc, -chapter_code, -days_to_readmission)
```

### Dropping deceased patients from the data frame
```{r}
rra_df = rra_df %>% filter(!(discharge_disposition == "expired"))
```

### Splitting the data frame into test and train sets

```{r}
set.seed(73)
sample_size <- floor(0.80 * nrow(rra_df))
train_indices <- sample(seq_len(nrow(rra_df)), size = sample_size)

train_df <- rra_df[train_indices, ]
test_df <- rra_df[-train_indices, ]
```

### Separating features (X) and target (y)
### Converting features to a numeric matrix (this also handles dummy encoding)

```{r}
X_train <- model.matrix(train_df$readmitted_within_30days ~ ., data = train_df)
y_train <- train_df$readmitted_within_30days

X_test <- model.matrix(test_df$readmitted_within_30days ~ ., data = test_df)
y_test <- test_df$readmitted_within_30days
```

### Converting to required xgb.DMatrix format

```{r}
dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = y_train)
dtest <- xgb.DMatrix(data = as.matrix(X_test), label = y_test)
```

### Set parameters for the training model

```{r}
params <- list(
booster = "gbtree",
objective = "binary:logistic",
eta = 0.3,
gamma = 0,
max_depth = 6,
min_child_weight = 1,
subsample = 1,
colsample_bytree =1,
eval_metric = "auc",
eval_metric = "aucpr",
eval_metric = "logloss"
)
```

### Using XGBoost's cross-validation function
### Let's calculate the best number of rounds for this model.

```{r}
xgbcv <- xgb.cv(
params = params,
data = dtrain,
nrounds = 100,
nfold = 5,
showsd = TRUE,
stratified = TRUE,
print_every_n = 10,
early_stopping_rounds = 20,
maximize = FALSE
)
```

### Best iteration = 23

### Training the model

```{r}
xgb1 <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 23,
  evals = list(val = dtest, train = dtrain),
  print_every_n = 10,
  early_stopping_rounds = 10,
  maximize = FALSE,
)
```

### Testing the model prediction on the test set

```{r}
preds <- predict(xgb1, dtest)
```

### Setting an output threshold for the predictions

```{r}
binary_preds <- ifelse(preds > 0.5, 1, 0)
```
### Calculating Accuracy

```{r}
accuracy <- mean(binary_preds == y_test)
print(paste("Test Accuracy:", accuracy))
```
```{r}
importance_matrix <- xgb.importance(
feature_names = colnames(dtrain),
model = xgb1
)
print(importance_matrix)
```
### Plotting the Importance Matrix

```{r}
imp_df = xgb.importance(model = xgb1) %>%
  as.data.frame() %>%
  head(10)
```
```{r}
clean_labels <- c(
"num_lab_procedures" = "# Lab Procedures",
"age" = "Age",
"num_medications" = "# Medications",
"time_in_hospital" = "Length of Stay",
"hospital_id" = "Hospital ID",
"has_comorbidity" = "Has Comorbidity",
"num_prior_admissions" = "# Prior Admissions",
"insurance_typemedicare" = "Insurance Type: Medicare",
"gendermale" = "Gender: Male",
"insurance_typeprivate" = "Insurance Type: Private"
)
```

```{r}
imp_df <- imp_df %>%
  mutate(
    CleanName = ifelse(Feature %in% names(clean_labels), clean_labels[Feature], Feature),
    CleanName = fct_reorder(CleanName, Gain)
  )
```

```{r}
ggplot(imp_df, aes(x = Gain, y = CleanName)) +
  geom_col(fill = "#76B7B2") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Feature Importance (Gain)",
    x = "Relative Contribution",
    y = NULL
) +
  theme(
    axis.text.y = element_text(face = "bold"),
    panel.grid.major.y = element_blank()
) 
```
  